Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-24 related.tex
quitting for the night on related.tex

Modeling time-series data for the purposes of prediction is an old
problem that dates back at least to Yule's 1927 invention of
autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
of prediction strategies have been developed for a wide variety of
prediction tasks.  The purpose of this paper is not to add a new
method to this arsenal, nor to assess or compare the effectiveness of
these existing methods.  Our goals are broader: (i) to empirically
quantify the predictive structure that is present in a real-valued
scalar time series and (ii) to explore how that inherent complexity
affects the performance of those existing prediction methods.  It
would, of course, be neither practical nor interesting to report
results for every existing forecast method; instead, we choose a
representative set, which are described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is also not a new problem.  Most solutions fall into
two categories: model-based error analysis and model-free information
analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This can proceed locally or globally.  The local version of
this type of analysis approximates error distributions for different
regions of a time-series model using local ensemble in-sample
forecasting.  [[Joshua, we probably want a quick definition of
    in-sample and out-of-sample.]] These distributions are then used
to estimate out-of-sample forecast errors [[in that area of]] the time
series model.  For example, Smith {\sl et al.} make in-sample
forecasts using ensembles around selected points in the time series in
order to predict the local predictability of a time series
reconstructed using delay-coordinate embedding~\cite{Smith199250}.
This approach can be used to show that different portions of a time
series can exhibit varying levels of local predictive uncertainty.  We
expand on this finding later in this paper with a time series that
exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent redundancy of a
time series~\cite{crutchfield2003} and thus cannot be used to draw
conclusions about predictive structure that may be usable by other
forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed {\sl post facto} from a
class of models, to determine which of those models was best.  When
building an autoregressive model, for example, it is common to verify
that the model error is normally distributed.  If it is not, that
indicates that there is linear structure in the time series that the
model-building process failed to use.  The problem with this approach
is that if the error is normally distributed it can not be said that
the practitioner is using the ideal forecasting schema---e.g., if
nonlinear structure is present that linear modeling is oblivious
to---but only if this class of model used all the structure it was
able to process.

A practice known as deterministic vs. stochastic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global error analysis. The basic idea is to
construct a series of local linear fits (single point) to global
linear fits (fitting all points known at that time), increasing the
number of points in the fit slowly. Then analyze how the average
out-of-sample error changes as a function of number of points in the
linear fits. The shape of such a graph indicates how much determinism
vs. stochasticity is present in a time series, see
\cite{Casdagli92dvsplots, weigend-book} for more information.

@@

[[I think we need a short paragraph about redundancy in here
    somewhere.  We keep referring to it as if the reader knows it...]]

>>>>

Modeling time-series data for the purposes of prediction is an old
problem that dates back at least to Yule's 1927 invention of
autoregression~\cite{Yule27}.  Since then hundreds, if not thousands,
of strategies have been developed for a wide variety of prediction
tasks.  The purpose of this paper is not to add a new weapon to this
arsenal, nor to assess or compare the effectiveness of existing
methods.  Our goals are broader: (i) to empirically quantify the
predictive structure that is present in a real-valued scalar time
series and (ii) to explore how that inherent complexity affects the
performance of those existing prediction methods.  It would, of
course, be neither practical nor interesting to report results for
every existing forecast method; instead, we choose a representative
set, which are described in Section~\ref{sec:model}.

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is also not a new problem.  Most solutions fall into
two categories: model-based error analysis and model-free information
analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This can proceed locally or globally.  The local version of
this type of analysis approximates error distributions for different
regions of a time-series model using local ensemble in-sample
forecasting.  {\color{red}[[Joshua, we probably want a quick
      definition of in-sample and out-of-sample.]]} These
distributions are then used to estimate out-of-sample forecast errors
{\color{red}[[in that area of]] [[means?]]} the time series model.
For example, Smith {\sl et al.}  make in-sample forecasts using
ensembles around selected points in the time series in order to
predict the local predictability of a time series reconstructed using
delay-coordinate embedding~\cite{Smith199250}.  This approach can be
used to show that different portions of a time series can exhibit
varying levels of local predictive uncertainty.  We expand on this
finding later in this paper with a time series that exhibits
interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent redundancy of a
time series~\cite{crutchfield2003} and thus cannot be used to draw
conclusions about predictive structure that may be usable by other
forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed {\sl post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to go back
and verify that the model error is normally distributed.  If it is
not, that indicates that there is structure in the time series that
the model-building process was unable to capture and use.
{\color{red} [[The problem with this approach is that if the error is
      normally distributed it can not be said that the practitioner is
      using the ideal forecasting schema---e.g., if nonlinear
      structure is present that linear modeling is oblivious to---but
      only if this class of model used all the structure it was able
      to process.]]  [[having trouble understanding this...]]}

A practice known as deterministic vs. stochastic
modeling~\cite{Casdagli92dvsplots, weigend-book} bridges the gap
between local and global error analysis {\color{red} [[just in
      model-based error analysis?  or also in model-free?]]}  The
basic idea is to construct a series of local linear fits (single
point) to global linear fits (fitting all points known at that time),
increasing the number of points in the fit slowly, and then analyze
how the average out-of-sample error changes as a function of number of
points in the linear fits. The shape of such a graph indicates how
much determinism vs. stochasticity is present in a time series.

@@

[[I think we need a short paragraph about redundancy in here somewhere
    --- with a bit more info than was in the intro where we first
    defined it.  We keep referring to it as if the reader knows
    it...]]
